---
title: "Weekly ARIMA"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Document Overview

When training the models, we only used one location, the Water Lab and the training data. This is because it had the most data (least NA's).

We fitted two main models, one model is just the straight seasonal trend, the other included discharge. We wanted to predict the minimum level of dissolved oxygen, but the max and the mean are reported for future modeling. Each model is saved as a .rds for reference, as it took a while to run the code.

With each model, we predicted the next week (this is from the test data), then retrained the model with the same parameters with the original data and the next week's data. This new model was then used to predict the next weeks data. We then repeated, retraining the model with the same parameters using the original data, and now the next two weeks' data, then predicted the next week (the third week). We repeated this for a whole year. The last year of the test data is reserved for after all the model testing as a final test.

## Load Data

```{r}
#load data and packages
library(dplyr)
library(forecast)
library(ggplot2)
library(tidyr)
library(patchwork)

set.seed(123)
load("C:/Users/emmak/Downloads/LRO_eco_forecasting/Data/Logan_River_Train.RData")

load("C:/Users/emmak/Downloads/LRO_eco_forecasting/Data/Logan_River_Test.RData")


#TRAINING, cleaning and using only the water lab location
Logan_River_train %>%
  group_by(source) %>%
  summarise(n_less = sum(DO < 10, na.rm = TRUE))

Logan_River_Clean <- Logan_River_train %>%
   mutate(across(where(is.numeric),
                ~ ifelse(. < 0, NA, .))) %>%
  filter(grepl("Water Lab", source))

summary(Logan_River_Clean)

daily_DO <- Logan_River_Clean %>%
  mutate(date = as.Date(date)) %>%
  group_by(date) %>%
  summarize(DO_min = min(DO, na.rm = TRUE),
            DO_mean = mean(DO, na.rm = TRUE),
            DO_max = max(DO, na.rm = TRUE),
            discharge_ave = mean(discharge, na.rm = TRUE)
            ) %>%
  filter(is.finite(DO_min),
         is.finite(DO_mean),
         is.finite(DO_max),
         is.finite(discharge_ave))

#TESTING, cleaning and using only the water lab locatino
Logan_River_test %>%
  group_by(source) %>%
  summarise(n_less = sum(DO < 10, na.rm = TRUE))

Logan_River_Clean_Test <- Logan_River_test %>%
   mutate(across(where(is.numeric),
                ~ ifelse(. < 0, NA, .))) %>%
  filter(grepl("Water Lab", source))

summary(Logan_River_Clean_Test)

daily_DO_Test <- Logan_River_Clean_Test %>%
  mutate(date = as.Date(date)) %>%
  group_by(date) %>%
  summarize(DO_min = min(DO, na.rm = TRUE),
            DO_mean = mean(DO, na.rm = TRUE),
            DO_max = max(DO, na.rm = TRUE),
            discharge_ave = mean(discharge, na.rm = TRUE)) %>%
  filter(is.finite(DO_min),
         is.finite(DO_mean),
         is.finite(DO_max),
         is.finite(discharge_ave))

summary(daily_DO_Test)


head(daily_DO_Test)

```

## Seasonal Model

This is the model only using seasonality, with no other factors.

```{r}
#MINIMUM
ggplot(daily_DO, aes(x = date, y = DO_min))+
  geom_point()

#Model Fitting
  #ARIMA_min <- auto.arima(ts(daily_DO$DO_min, frequency = 365), seasonal = TRUE)
  #summary(ARIMA_min)
  #saveRDS(ARIMA_min, file = "ARIMA_min.rds")


#MEAN, this model is fit, but not used in testing
ggplot(daily_DO, aes(x = date, y = DO_mean))+
  geom_point()

#Model Fitting
  #ARIMA_mean <- auto.arima(ts(daily_DO$DO_mean, frequency = 365), seasonal = TRUE)
  #summary(ARIMA_mean)
  #saveRDS(ARIMA_mean, file = "ARIMA_mean.rds")

#mean one week prediction
forecast_week_mean <- forecast(ARIMA_mean, h = 7)
plot(forecast_week_mean)

```

## Predicting using Seasonal Model with RMSE

```{r}
#PREDICT

forecast_min_weekly <- list()

for(week in 1:52){
    new_data <- daily_DO_Test$DO_min[seq_len((week-1)*7)]

    combined <- c(daily_DO$DO_min, new_data)
    
    ts_data <- ts(combined, start = c(2014,1), frequency = 365)

    temp_fit <- Arima(ts_data, model = ARIMA_min)

    predict_seven <- predict(temp_fit, n.ahead = 7)
    
    forecast_min_weekly[[week]] <- data.frame(
                             pred = predict_seven$pred,
                             se = predict_seven$se,
                             week = week,
                             forecast_day = seq(1:7)
    )
}

forecast_min_weekly_df <- bind_rows(forecast_min_weekly)
summary(forecast_min_weekly_df)

forecast_min_weekly_df <- forecast_min_weekly_df %>%
  mutate(day = 1:n())

ggplot(data = forecast_min_weekly_df, aes(x = day, y = pred))+
  geom_line()
```

```{r}
#RMSE of day of the week

#Forecast Day Averages
forecast_day_ave <- forecast_min_weekly_df %>%
  mutate(
    obs = daily_DO_Test$DO_min[1:364],
    week = rep(1:52, each = 7),
    day = rep(1:7, times = 52)
  ) %>%
  group_by(day) %>%
  summarise(
    rmse = sqrt(mean((obs - pred)^2, na.rm = TRUE))
  )


#graph
ggplot(forecast_day_ave, aes(x = day, y = rmse))+
  geom_point()
```

## Seasonal model with discharge: Auto ArimaX

```{r}
#Add lag (using previous week's discharge level)
lag_df <- daily_DO %>%
  mutate(discharge_lag = lag(discharge_ave, 7)) %>%
  filter(!is.na(discharge_lag))

#Model
  #ARIMAX_min <- auto.arima(ts(lag_df$DO_min, frequency = 365), xreg =       lag_df$discharge_lag,  seasonal = TRUE)
  #saveRDS(ARIMAX_min, file = "ARIMAX_min.rds")
summary(ARIMAX_min)

#Forecasting
forx_min_weekly <- list()

for(week in 1:52){
    new_data <- daily_DO_Test$DO_min[seq_len((week-1)*7)]
    combined <- c(lag_df$DO_min, new_data)
    
    newx <- lag(daily_DO_Test$discharge_ave, 7)[seq_len((week - 1) * 7)]
    combinedx <- c(lag_df$discharge_lag, newx)
    
    ts_data <- ts(combined, start = c(2014,1), frequency = 365)
    temp_fit <- Arima(ts_data, model = ARIMAX_min, xreg = combinedx)
    
    full_discharge <- c(daily_DO$discharge_ave,
                      daily_DO_Test$discharge_ave[seq_len((week - 1) * 7)])
  
    future_xreg <- tail(full_discharge, 7)

    predict_seven <- predict(temp_fit, n.ahead = 7, newxreg = future_xreg)
    
    forx_min_weekly[[week]] <- data.frame(
                             pred = predict_seven$pred,
                             se = predict_seven$se,
                             week = week,
                             forecast_day = seq(1:7)
    )
}

forx_min_weekly_df <- bind_rows(forx_min_weekly)
summary(forx_min_weekly_df)

forx_min_weekly_df <- forx_min_weekly_df %>%
  mutate(day = 1:n())

ggplot(data = forx_min_weekly_df, aes(x = day, y = pred))+
  geom_line()

```

```{r}
#RMSE

#Forecast Day Averages
forx_day_ave <- forx_min_weekly_df %>%
  mutate(
    obs = daily_DO_Test$DO_min[1:364],
    week = rep(1:52, each = 7),
    day = rep(1:7, times = 52)
  ) %>%
  group_by(day) %>%
  summarise(
    rmse = sqrt(mean((obs - pred)^2, na.rm = TRUE))
  )

#graph
ggplot(forx_day_ave, aes(x = day, y = rmse))+
  geom_point()
```

```{r}

```

## Comparing Model Predictions to Actual (with CI)

```{r}

#Auto Arima
forecast_min_weekly_df$obs <- daily_DO_Test$DO_min[1:364]
forecast_min_weekly_df <- forecast_min_weekly_df %>%
  mutate(
    lower = pred - 1.96 * se,
    upper = pred + 1.96 * se
  )

ggplot(forecast_min_weekly_df, aes(x = day))+
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, fill = "lightblue") +
  geom_line(aes(x = day, y = obs), col = "black")+
  geom_line(aes(x = day, y = pred), col = "steelblue")


#Arimax
forx_min_weekly_df$obs <- daily_DO_Test$DO_min[1:364]
forx_min_weekly_df <- forx_min_weekly_df %>%
  mutate(
    lower = pred - 1.96 * se,
    upper = pred + 1.96 * se
  )

ggplot(forx_min_weekly_df, aes(x = day))+
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, fill = "pink") +
  geom_line(aes(x = day, y = obs), col = "black")+
  geom_line(aes(x = day, y = pred), col = "red3")
  

```

## Comparing Model predictions between models and Actual

This uses a rolling average to smooth the lines a bit so we can actually see what is happening and compare without too much noise.

```{r}
#Atuo Arima Rolling Averages
forecast_min_weekly_df$pred_smooth <- as.numeric(
  stats::filter(
    forecast_min_weekly_df$pred,
    rep(1/7, 7),
    sides = 1
  )
)

forecast_min_weekly_df$obs_smooth <- as.numeric(
  stats::filter(
    forecast_min_weekly_df$obs,
    rep(1/7, 7),
    sides = 1
  )
)

forecast_min_weekly_df$upper_smooth <- as.numeric(
  stats::filter(
    forecast_min_weekly_df$upper,
    rep(1/7, 7),
    sides = 1
  )
)

forecast_min_weekly_df$lower_smooth <- as.numeric(
  stats::filter(
    forecast_min_weekly_df$lower,
    rep(1/7, 7),
    sides = 1
  )
)

ggplot(forecast_min_weekly_df, aes(x = day))+
  geom_ribbon(aes(ymin = lower_smooth, ymax = upper_smooth), alpha = 0.2, fill = "lightblue") +
  geom_line(aes(x = day, y = obs_smooth), col = "black")+
  geom_line(aes(x = day, y = pred_smooth), col = "steelblue")

#Arimax  Rolling Averages
forx_min_weekly_df$pred_smooth <- as.numeric(
  stats::filter(
    forx_min_weekly_df$pred,
    rep(1/7, 7),
    sides = 1
  )
)

forx_min_weekly_df$obs_smooth <- as.numeric(
  stats::filter(
    forx_min_weekly_df$obs,
    rep(1/7, 7),
    sides = 1
  )
)

forx_min_weekly_df$upper_smooth <- as.numeric(
  stats::filter(
    forx_min_weekly_df$upper,
    rep(1/7, 7),
    sides = 1
  )
)

forx_min_weekly_df$lower_smooth <- as.numeric(
  stats::filter(
    forx_min_weekly_df$lower,
    rep(1/7, 7),
    sides = 1
  )
)

ggplot(forx_min_weekly_df, aes(x = day))+
  geom_ribbon(aes(ymin = lower_smooth, ymax = upper_smooth), alpha = 0.2, fill = "pink") +
  geom_line(aes(x = day, y = obs_smooth), col = "black")+
  geom_line(aes(x = day, y = pred_smooth), col = "red3")


#Combined Graphs of both Models
forx_min_weekly_df$model <- "Includes Discharge"
forecast_min_weekly_df$model <- "Excludes Discharge"

g1 <- ggplot() +
  geom_ribbon(aes(x = day, ymin = lower_smooth, ymax = upper_smooth, 
                  fill = model), 
              data = forx_min_weekly_df, alpha = 0.2) +
  geom_ribbon(aes(x = day, ymin = lower_smooth, ymax = upper_smooth, 
                  fill = model), 
              data = forecast_min_weekly_df, alpha = 0.2) +
  geom_line(aes(x = day, y = pred, color = model), 
            data = forx_min_weekly_df) +
  geom_line(aes(x = day, y = pred, color = model), 
            data = forecast_min_weekly_df) +
  geom_line(aes(x = day, y = obs, color = "Observed"), 
            data = forx_min_weekly_df) +
  scale_fill_manual(values = c("Includes Discharge" = "red3", 
                               "Excludes Discharge" = "steelblue")) +
  scale_color_manual(values = c("Includes Discharge" = "red3", 
                                "Excludes Discharge" = "steelblue",
                                "Observed" = "black"
                                )) +
  labs(color = "Model", fill = "Confidence Interval", 
       y = "Minimum DO (mg/L)", x = "Day of Year",
       title = "Forecasting a Year From 2/15/2024 ")
g1



```

## Discharge Graph over time

```{r}
dishcarge_df <- data.frame(
  day = 1:364,
  discharge = daily_DO_Test$discharge_ave[1:364]
)

g2 <- ggplot(dishcarge_df, aes(x = day, y = discharge))+
  geom_line()+
  labs(x = "Day of Year",
       y = "Discharge (m^3/s)")
```

## RMSE over time

```{r}

library(slider)

#RMSE with no discharge
noq_rmse <- slide_dbl(
  (forecast_min_weekly_df$obs - forecast_min_weekly_df$pred)^2,
  ~ sqrt(mean(.x)),
  .before = 7 - 1,
  .complete = TRUE
)

#rmse with discharge
q_rmse <- slide_dbl(
  (forx_min_weekly_df$obs - forx_min_weekly_df$pred)^2,
  ~ sqrt(mean(.x)),
  .before = 7 - 1,
  .complete = TRUE
)

rmse_df <- data.frame(
  noq_rmse = noq_rmse,
  q_rmse = q_rmse,
  day = seq(1:364)
)

g3 <- ggplot(rmse_df, aes(x = day)) +
  geom_line(aes(y = noq_rmse, color = "Excludes Discharge")) +
  geom_line(aes(y = q_rmse, color = "Includes Discharge")) +
  scale_color_manual(values = c("Excludes Discharge" = "steelblue",
                                "Includes Discharge" = "red3"
                                )) +
  labs(color = "Model", y = "RMSE", x = "Day of Year")+
  theme(legend.position = "none")
```

## Combine previous three graphs to compare over time

Final graph is located under "Figures" called "Stack of Graphs"

```{r}
g1 / g2 / g3 + plot_layout(
  axis_titles = "collect",
  axes = "collect"
)
```

## Compare Oxygen to Discharge With Time of Year

Final graph is located under "Figures" called "Discharge and Min DO Comparison"

```{r}
#Getting day of year not as a date
lag_df$day <- as.integer(format(lag_df$date, "%j"))

#Plotting relationship of Lagged Discharge to Dissolved Oxygen Level
ggplot(lag_df, aes(x = discharge_lag, y = DO_min, col = day))+
  geom_point()+
  scale_color_gradient(low = "#fdb863",
                       high = "#542788")+
  labs(color = "Day of Year", 
       x = "7-Day Lagged Discharge Level",
       y = "Minimum DO (mg/L)",
       title = "Discharge and Minimum DO Comparison")
```

## Looking at RMSE with high Q and low Q

Final graph is located under "Figures" called "Average RMSE of Week High Q v Low Q"

```{r}
df_combined <- forecast_min_weekly_df %>%
  left_join(lag_df %>% select(day, discharge_lag), by = "day")  

rmse_by_day <- df_combined %>%
  mutate(
    discharge_level = ifelse(discharge_lag > 10, "High", "Low"), #High discharge is above 10
    day_of_week = forecast_day,
    error_sq = (obs - pred)^2
  ) %>%
  group_by(discharge_level, day_of_week) %>%
  summarise(rmse = sqrt(mean(error_sq, na.rm = TRUE)))


#Plot
ggplot(rmse_by_day, aes(x = day_of_week, y = rmse, col = discharge_level)) +
  geom_point(cex = 3) +
  scale_color_manual(values = c("High" = "#1b9e77", 
                                "Low" = "#d95f02"))+
  labs(color = "Discharge Level", x = "Forecast Day", y = "Average RMSE",
       title = "Averaged RMSE Over Forecast Week")+
  scale_x_continuous(breaks = seq(1,7, by = 1))



```
